\chosentitle

### 2024-08-21 paper/main.tex
update methods

\documentclass[modern]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}

@@

\newcommand{\project}[1]{\textit{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vectheta}{\boldsymbol{\theta}}
\newcommand{\vecpsi}{\boldsymbol{\psi}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecX}{\mathbf{X}}
\newcommand{\hadamard}{\odot}
\newcommand{\apogee}{\project{APOGEE}}
\newcommand{\boss}{\project{BOSS}}
\newcommand{\sdss}{\project{SDSS}}

@@

\begin{abstract}\noindent
Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set. 
These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
Here we present a \emph{linear} model that simultaneously fits stellar absorption and the joint continuum-instrument response.
The joint continuum-instrument response is modelled with a Fourier basis, and stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors. 
The non-negativity constraint ensures that basis vectors are strictly additive.
Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
We demonstrate this method by simultaneous fits to high-resolution echelle spectra, where continuum is usually fit per order.
When we restructure our forward model to \emph{linearly} interpolate from stellar parameters to basis weights, we show the the stellar parameters we infer from \project{Sloan Digital Sky Survey} infrared spectra are in closer agreement with \todo{astrophysical expectations} than those reported using neural networks or high-dimensional cubic spline interpolation.

%Continuum normalization is often a necessary step when analyzing stellar spectra. 
%The ideal approach might be to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is often required before stellar parameters are estimated.
%Here we present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization that models stellar line absorption, and a sines-and-cosines basis to model the joint continuum-instrument response.
%The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
%The linearity of both components ensures that inference is stable and fast.
%Our approach factorizes line absorption entirely from continuum-normalized theoretical spectra, but we also describe a data-driven variant to iteratively correct model-data mismatches.
%We show that the basis weights can also be used to reliably estimate stellar parameters, and apply these methods to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type stars.
%This approach reduces subjectivity in estimating stellar continuum.
\end{abstract}

@@

In an ideal scenario the continuum might be fit jointly with the stellar parameters, but this is often too expensive if it is costly to predict the emergent spectrum. In Section~\ref{sec:methods} we describe a family of methods to address these problems. 
The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption. NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements. This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques. 
%The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \project{Sloan Digital Sky Survey} (\sdss). 
\todo{We apply these to SDSS.}
We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\

@@

Here we will describe the method in general before outlining explicit assumptions and implementation details. The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for these data can be expressed as the element-wise multiplication of what we will describe as the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response model $g(\lambda_i;\vectheta)$
\begin{align}
    y &= f(\lambda;\vecpsi)\hadamard{}g(\lambda;\vectheta) + \mbox{noise}
\end{align}
where the components $f(...)$ and $g(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ where the covariance matrix $\vec{C}$ has the transformed flux variances $\vec{C}_{ii} = (y_{i}\sigma_{y,i})^2$ along the diagonal. The log-transformation changes the element-wise (Hadamard; $\hadamard$) product into the summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecpsi)} + \log{g(\vec{\lambda};\vectheta)} \quad .
\end{align}

The line absorption model $f(\lambda_i;\vecpsi)$ predicts the line absorption at wavelength $\lambda_i$ given parameters $\vecpsi$. We construct the line absorption model $f(\lambda_i;\vecpsi)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF). In the applications that we present here, the theoretical spectra used to approximate the line absorption model are assumed to have the same wavelength sampling and line spread function as the data, such that we use a grid of model spectra where $D = P$. Other situations are permitted, but at inference time there is a need to evaluate the line absorption model to the $P$ observed wavelengths.\\

We refer to our $N \times D$ matrix of continuum-rectified theoretical spectra as $\vec{M}$. This is a dense matrix: there are no entries of exactly zero. However, a small transformation
%\begin{eqnarray}
%    \vec{X} \equiv -\log\left({\vec{M}}\right)
%\end{eqnarray}
to this matrix can make it it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{M}$ into two smaller matrices $\vec{W}$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{M}}\right) \approx \vec{W}\transpose\vec{F}
\end{eqnarray}
where $\vec{W}$ is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. All elements in $-\log\left({\vec{M}}\right)$, $\vec{W}$, and $\vec{F}$ are required to be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum. Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\

@@

%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecpsi$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.

The factorization of $-\log\left({\vec{M}}\right)$ into $\vec{W}\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{M}}\right)$: this is the default behaviour in many packages. We found that factorizing $\vec{W}\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex and do not require many choices.\\
%At inference time, $\vecpsi$ is analogous to the basis weights found at training time \\

@@

\noindent{}With the matrix $\vec{F}$ we can now define the logarithm of line absorption function $\log{f(\lambda_i;\vecpsi)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
\begin{align}
    \log{f(\lambda_i;\vecpsi)} = -\vec{F}\transpose\vecpsi \label{eq:f}
\end{align}
where $\vecpsi \in [0, \infty)$ is a column vector of $C$ \emph{basis weights} to be solved at inference time. Here, $\vecpsi$ is analogous to a column in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors $\vec{F}$. Note that because $\vecpsi$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response. \\

There are many suitable choices for the logarithm of the continuum-instrument response model $\log{g(\lambda_i;\vectheta)}$. Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $\log{g(\lambda_i;\vectheta)}$ is expressed compactly as
\begin{align}
    \log{g(\lambda_i;\vectheta)} = \vec{G}\vectheta
\end{align}
where $\vec{G}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
\begin{align}
    \vec{G}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}
\noindent{}and the design matrix $\vec{G}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins. Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\

\noindent{}With $f(\lambda_i;\vecpsi)$ and $g(\lambda_i;\vectheta)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecpsi + \vec{G}\vectheta
\end{equation}
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\\vec{G}\end{bmatrix} \quad .
    \quad \mbox{and} \quad 
    \vec{X} = \begin{bmatrix}\vecpsi\\\vectheta\end{bmatrix} 
\end{eqnarray}
Given some observed log-transformed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem which is subject to the constraint $\vecpsi \geq 0$
\begin{eqnarray}
    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad ,
\end{eqnarray}

\noindent{}where the un-transformed rectified flux and continuum can be computed in a straightorward manner:
\begin{eqnarray}
    f(\boldsymbol{\tilde{\psi}}) &\,=\,& \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\psi}}\right)} \\
    g(\boldsymbol{\tilde{\theta}}) &\,=\,& \exp{\left(\vec{G}\boldsymbol{\tilde{\theta}}\right)}
\end{eqnarray}

@@

%The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed. We experimented with choices of initialisation and inference. Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously. The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip). Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios. \\

@@

% TODO: re-word this
\section{Data}

- SDSS
%The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO). Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper (MWM) `carton'\footnote{`Cartons' are how targets are now assigned in SDSS.}, or because they are standard stars. We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton. \todo{Some description of stellar-like.}\\

@@

%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecpsi$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data. We found that setting $\vecpsi$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization. In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecpsi$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.

@@

We applied the three variants to all \boss\ stellar spectra currently available in the \emph{Sloan Digital Sky Survey}. This includes approximately \todo{X} spectra of \todo{Y} stars.

\begin{figure}
    \caption{Continuum-normalized \boss\ spectra of all spectral types, with the best-fitting model spectrum shown.}
\end{figure}

\begin{figure}
    \caption{A Gaia H-R diagram showing the median $\chi^2$ per bin.}
\end{figure}

@@

\begin{figure}
    \caption{Gaia H-R diagram with each bin coloured by the median $\vecpsi_i$ value, picking some $i$th eigenspectra that looks like metallicity. \label{fig:gaia_hrd_metallicity}}
\end{figure}

\begin{figure*}
    \caption{The median pixel $\chi^2$ value as a function of de-reddened \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence [OR GIANT?] stars. \todo{Should expect to see increasing residuals due to features not captured by the model, and emission lines.}}
    % Could do the same showing standard deviation per bin
\end{figure*}

@@

- getting good fits even with bad models: the hydrogen lines in the apogee grid are knowingly incorrect, but we see no difference in the $\chi^2$ at those wavelengths. That is because the hydrogen lines separate out into very clean eigenspectra, where the amplitude for that eigenspectra is not required to vary smoothly with other amplitudes

- biases as a function of S/N or labels

@@

- $\chi^2$ across the HRD

@@

- issue with M-giant eigenspectra when they were over-represented in the sample

@@

We solved continuum normalization.

@@

\paragraph{Acknowledgements}
It is a pleasure to thank
% All these people are likely co-authors, but should be thanked if they don't want to be co-authors
    Adam Wheeler (Ohio State University),
    Andrew Saydjari (Harvard),
    David W. Hogg (New York University),
    Megan Bedell (Flatiron Institute),
    Michael Blanton (NYU)
.
% include bibliography
\bibliographystyle{aasjournal}
%\bibliography{bibliography}

>>>>

\documentclass[modern]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{MnSymbol}

@@

\newcommand{\project}[1]{\textit{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vectheta}{\boldsymbol{\theta}}
\newcommand{\vecalpha}{\boldsymbol{\alpha}}
\newcommand{\vecbeta}{\boldsymbol{\beta}}
\newcommand{\vecgamma}{\boldsymbol{\gamma}}
\newcommand{\vecW}{\mathbf{W}} % stellar line absorption basis weights
\newcommand{\vecF}{\mathbf{F}} % stellar line absorption basis vectors
\newcommand{\vecG}{\mathbf{G}} % telluric line absorption basis vectors
\newcommand{\vecH}{\mathbf{H}} % continuum basis vectors
\newcommand{\vecX}{\mathbf{X}}

\newcommand{\hadamard}{\odot}
\newcommand{\apogee}{\project{APOGEE}}
\newcommand{\boss}{\project{BOSS}}
\newcommand{\sdss}{\project{SDSS}}
\newcommand{\eso}{\project{ESO}}
\newcommand{\harps}{\project{HARPS}}

@@

\begin{abstract}\noindent
Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set. 
These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
Here we present a \emph{linear} model that simultaneously fits stellar absorption and the joint continuum-instrument response.
The joint continuum-instrument response is modelled with a Fourier basis, and stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors. 
We factorize telluric absorption in the same way, allowing us to simultaneously fit stellar and telluric absorption.
The non-negativity constraint ensures that basis vectors are strictly additive.
Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
This model allows us to reliably fit the stellar absorption, telluric absorption, the continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
We demonstrate this method by fits to high-resolution echelle spectra, where we achieve \todo{X\%} accuracy in continuum normalization with no apparent biases at low signal-to-noise ratios (e.g., $<10$).
\end{abstract}

@@

In an ideal scenario the continuum might be fit jointly with the stellar parameters, but this is often too expensive if it is costly to predict the emergent spectrum. In Section~\ref{sec:methods} we describe a family of methods to address these problems. 
The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption. NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements. This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques. 
We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\

@@

Here we will describe the method in general before outlining explicit assumptions and implementation details. The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric absorption model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
\begin{align}
    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
\end{align}
where the components $f(...)$, $g(...)$, and $h(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
\begin{eqnarray}
    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
\end{eqnarray}
\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into the summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
\end{align}\\

We now turn to defining the stellar absorption model, the telluric absorption model, and the continuum model.
The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$. We construct the line absorption model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF). 
The theoretical spectra used to construct  the line absorption model do not have to have the same wavelength sampling and instrument line spread profile as the data, but at inference time there is a need to interpolate (or evaluate) the line absorption model to the $P$ observed wavelengths.\\

We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$. This is a dense matrix: there are no entries of exactly zero. However, a small transformation to this matrix can make it it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
\end{eqnarray}
where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. All elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ are required to be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\

@@

%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecalpha$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.

The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages. We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex, do not require many choices, and they make for an efficient problem for graphical processing units.\\
%At inference time, $\vecalpha$ is analogous to the basis weights found at training time \\

\todo{We used the BOSZ grid. Some details about that, including settings used for factorizing.}\\

@@

\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line absorption function $\log{f(\lambda_i;\vecalpha)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
\begin{align}
    \log{f(\lambda_i;\vecalpha)} = -\vec{F}\transpose\vecalpha \label{eq:log-f}
\end{align}
where $\vecalpha \in [0, \infty)$ is a column vector of $K$ \emph{stellar basis weights} to be solved at inference time. Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the stellar absorption from the $K$ basis vectors $\vec{F}$. Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response. \\

Telluric line absorption can be included in a similar way to the stellar line absorption. We factorize the negative log of theoretical telluric absorption into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$.
\todo{Details about what telluric grid we used, and setting used for factorizing. Note how many vectors we chose ($L$)} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
\begin{eqnarray}
    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
\end{eqnarray}
where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} to be solved at inference time. We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}). Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\

There are many suitable choices for the logarithm of the continuum-instrument response model $\log{h(\lambda_i;\vecgamma)}$. Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $\log{h(\lambda_i;\vecgamma)}$ is expressed compactly as
\begin{align}
    \log{h(\lambda_i;\vecgamma)} = \vec{G}\vecgamma
\end{align}
where $\vec{H}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
\begin{align}
    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}
\noindent{}where $L$ \todo{overloaded} is a length scale. \todo{Talk more about this.}
The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins. Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\

\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
\end{equation}
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
    \quad \mbox{and} \quad 
    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
\end{eqnarray}

Given some log-transformed observed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem. This could be solved directly by linear algebra,
\begin{eqnarray}
    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
\end{eqnarray}
\noindent{}but this does not guarantee that the bounds $\vecalpha \geq 0$ and $\vecbeta \geq 0$ are met.

\todo{Talk about how we can solve this as} $||\vec{Y}/\sqrt{C} - \vec{AX}\sqrt{C}||^2$ with linear operator, truncated reflective region, or bounded least-squares with a provided jacobian ($\vec{A}$).  Thankfully, each of these optimization paths is convex.\\

\todo{Move this to 3.2?}
In later sections we describe some situations where we may want to penalize values of $\vecalpha$. We perform this by feature weighting, which is equivalent to setting a prior on the distribution that $\vecalpha$ can be drawn from. We introduce the feature weighting matrix $\vec{\Lambda}$, where $K$ is the number of basis vectors chosen for NMF and $L$ is the number of Fourier modes. We set the first $K \times K$ elements of $\vec{\Lambda}$ to $\mathrm{Cov}{\left(\vec{W}\vec{W}\transpose\right)}$ and set all other elements of $\vec{\Lambda}$ to zero. We can scale the strength of this feature weighting by a scalar $\lambda$ (\todo{nomenclature overload of lambda}) and solve for the parameters $\vec{\tilde{X}}$ with
\begin{eqnarray}
    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad .
\end{eqnarray}\\

After solving for $\vec{\tilde{X}}$, the (un-transformed) rectified flux and continuum can be computed in a straightorward manner:
\begin{eqnarray}
    \mathrm{rectified~stellar~flux}, &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
    \mathrm{telluric~absorption}, &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
    \mathrm{continuum}, &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
\end{eqnarray}

@@

%The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed. We experimented with choices of initialisation and inference. Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecalpha$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecalpha,\vectheta\}$ simultaneously. The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecalpha$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip). Initializing from small ($10^{-12}$) values of $\vecalpha$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecalpha$) seemed to work well in many scenarios. \\

@@

% TODO: re-word this

\section{Data}

We demonstrate the effectiveness of our method by forward modeling high-resolution optical echelle spectra taken with the MIKE spectrograph on the Magellan Clay telescope. These spectra have a spectral resolution of $\mathcal{R} \approx X$ and a wavelength range from $X$ to $Y$\,nm. The spectra are recorded on the 2D image plane as a series of $X$ orders: $X$ in the blue arm (X to Y) and $Y$ in the red arm (Y to Z). \\

\todo{Description of the blaze function, although this is not known perfectly}

The typical analysis procedure for this kind of spectra might involve selecting (by hand) pixels that do not appear to be obviously affected by line (or molecular) absorption, and fitting a smooth function to represent the continuum. This process would be repeated per-order, often ignoring the selection of pixels in the previous order (which overlaps in wavelength). A subsequent local continuum determination would be made for each line.

\todo{Details about the synthetic grid used, and how we trained the NMF. Include details about the telluric basis used as well.}\\

\todo{Assumptions about the continuum parameters}\\

We measure the radial velocity by cross-correlating the spectra against all basis vectors. \todo{More details about this.} In practice this alternating

\subsection{High-resolution infrared spectra}
- SDSS
%The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO). Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper (MWM) `carton'\footnote{`Cartons' are how targets are now assigned in SDSS.}, or because they are standard stars. We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton. \todo{Some description of stellar-like.}\\

@@

%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecalpha$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data. We found that setting $\vecalpha$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization. In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecalpha$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.

@@

We want to show:
\begin{enumerate}
    \item How does the mean/median rectified flux vary as a function of S/N for the same star (e.g., 18 Sco)?
    \item What do the fits look like for a few carefully chosen spectral types?
    \item What regularisation strength did we need to apply?
\end{enumerate}

@@

@@

- getting good fits with bad models

- model interpretability

- biases as a function of S/N

@@

@@

- setting the right covariance strength for a given model.

@@

@@

\paragraph{Acknowledgements}
It is a pleasure to thank
% All these people are likely co-authors, but should be thanked if they don't want to be co-authors
    Rory Smith (Monash),
    Adam Wheeler (Ohio State University),
    Andrew Saydjari (Harvard),
    David W. Hogg (New York University),
    Megan Bedell (Flatiron Institute),
    Michael Blanton (NYU)
.
% include bibliography
\bibliographystyle{aasjournal}
%\bibliography{bibliography}
